\newpage

\section{Initial Network Configuration}

This section describes the step-by-step laboratory environment setup process, including GNS3 setup, router configuration, and Docker container deployment. The configuration creates the underlying network infrastructure to be utilized by the VPN implementations.

\subsection{GNS3 Environment Setup}

Here we detail the technical steps required for implementing the network topology presented in Section 3. All device connections and configuration follow architecture specs provided in the GNS3 Network Simulation Platform subsection.

\subsubsection{Device Import and Template Creation}

\textbf{Cisco Router Template:}

\noindent
We have chosen for this activity the Cisco 7200 router running IOS version 124-24.T5. This specific IOS image is chosen because it is a freely downloadable equivalent of modern commercial Cisco IOS releases. Whilst newer platforms like IOSv (IOS Virtual) are faster and more feature-rich, they require valid Cisco license agreements that are too costly for learning purposes. The 124-24.T5 image, although older and not usually found in modern production settings, has all the fundamental routing and NAT functionality for this lab exercise. To import the router template:

\begin{enumerate}
    \item Open GNS3 and click the \textbf{router icon} in the left panel
    \item Select \textbf{"New Template"} at the bottom
    \item Choose \textbf{"Install an appliance from the GNS3 server"}
    \item Filter by \textbf{"CISCO"} and locate \textbf{"Cisco 7200"}
    \item Install the template and select version \textbf{"124-24.T5"}
    \item Load the router image file to complete the setup
\end{enumerate}

\noindent
\textbf{Docker Container Setup:}

\noindent
Two Docker containers are required for the VPN endpoints \cite{docker_official}:

\begin{itemize}
    \item \textbf{Server Container Image:} \texttt{siomiz/softethervpn:latest} \cite{siomiz_softether}
    \item \textbf{Client Container Image:} \texttt{ubuntu:latest} \cite{ubuntu_docker}
\end{itemize}

\noindent
To insert containers in GNS3 project:

\begin{enumerate}
    \item Navigate to \textbf{Edit → Preferences} in GNS3
    \item Select \textbf{"Docker containers"} from the left panel
    \item Click \textbf{"New"} to add a container
    \item Enter the image name (e.g., \texttt{siomiz/softethervpn:latest})
    \item Configure the container name and complete the setup
    \item Do the same for the Ubuntu client container
\end{enumerate}

\noindent
You should be able to see the containers in the left panel of GNS3 now, ready to be pulled into the workspace. So now you can configure the same topology as shown in Figure \ref{fig:gns3_topology} in Section 3 of this document. 

\noindent
Clicking on the wire on the left panel, you can connect the devices to each other, creating the network topology described. Choose the correct interfaces for each link as shown in the table in Section 3.2. The interfaces connected from the Client router and Server one to the internal nodes, are always the FastEthernet0/0, instead the FastEthernet0/1 one is the one connected to the ISP. The ISP with FastEthernet0/0 is connected to the server router, 0/1 is connected instead to the client router. There is, here besides, another interface, the one of FastEthernet1/0, which will be instead plugged into the internet bridge, so with the cloud and directly to the right interface of the host machine that provides the internet connection.

\noindent
\textbf{Critical Router Configuration Note:}
\noindent
Before proceeding with the router configurations, you must configure the ISP router to have three interfaces instead of the default two. This is needed because the ISP router must connect to both edge routers and the Internet cloud. To perform the ISP router configuration:

\begin{enumerate}
    \item Right-click on the ISP router in GNS3
    \item Select \textbf{"Configure"} from the context menu
    \item Navigate to the \textbf{"Slots"} section in the configuration dialog
    \item Replace the existing slot configuration from \textbf{"C7200-IO-FE"} to \textbf{"C7200-IO-2FE"}
    \item Apply the changes and close the configuration dialog
\end{enumerate}

This configuration change provides the ISP router with three FastEthernet interfaces (Fa0/0, Fa0/1, and Fa1/0) as required by the network topology, allowing it to connect to both edge routers and the Internet cloud simultaneously.

\subsection{Container Configuration}

The Docker containers must have special configuration for persistent network and storage setup.

\subsubsection{Server Container Configuration}

SoftEther VPN server container needs permanent directories to store configuration files:

\begin{enumerate}
    \item Open the server container configuration in GNS3 (right-click on the container)
    \item Navigate to \textbf{Advanced Settings}
    \item Add the following additional directory:
    \begin{itemize}
        \item \texttt{/usr/vpnserver}
    \end{itemize}
\end{enumerate}

\noindent
This directory will be created in the GNS3 project directory on the local system, which will allow persistent storage of configuration files during restart of the container.

\noindent
\textbf{Server Configuration File Setup:}

\noindent
In order to ensure the SoftEther VPN server starts with the proper configuration, you should place the \texttt{vpn\_server.config} file in the correct persistent directory:

\begin{enumerate}
    \item Navigate to your local file system GNS3 project directory 
    \item Open the \texttt{project-files} folder
    \item Look for the \texttt{docker} subdirectory
    \item Find the server container folder by means of the container ID
    \item Enter into the \texttt{usr/vpnserver} directory within the container folder
    \item Place the \texttt{vpn\_server.config} file there
\end{enumerate}

\noindent
This will enable the pre-configured settings, like user accounts, IPSec setup, and SSL/TLS configuration, to be loaded automatically when the container starts as started by the SoftEther VPN service.

\subsubsection{Client Container Setup}

The client container requires similar setup for persistent storage:

\begin{enumerate}
    \item Open the client container configuration in GNS3
    \item Navigate to Advanced Settings  
    \item Add the additional directory: \texttt{/client}
\end{enumerate}

\noindent
\textbf{Client Configuration Files Setup:}

\noindent
The client container requires a number of configuration files for IPSec and TLS VPN implementations. Place all the following files in the client's persistent directory:

\begin{enumerate}
    \item Navigate to your GNS3 project directory on the local filesystem
    \item Open the \texttt{project-files} folder
    \item Locate the \texttt{docker} subdirectory
    \item Find the client container folder
    \item Navigate to the \texttt{client} directory within the container folder
    \item Place the following files in this directory:
    \begin{itemize}
        \item \textbf{IPSec files:} \texttt{ipsec.conf} and \texttt{ipsec.secrets}
        \item \textbf{TLS/OpenVPN files:} \texttt{softether.ovpn}, \texttt{ca.crt}, and \texttt{credentials.txt}
    \end{itemize}
\end{enumerate}

\noindent
This setup ensures that all VPN client configuration setups are available when the container boots in order to create smooth connections with both IPSec and TLS VPN servers.

\subsection{Project Startup and Device Access}

After each of the configurations is completed and files are properly placed inside their persistent folders, you can initiate the lab environment.

\subsubsection{Launching the GNS3 Project}

To launch all devices and containers in the project:

\begin{enumerate}
    \item Click the green \textbf{Play button} in the top toolbar of GNS3
    \item This will boot every Docker container and router simultaneously
    \item Wait until all the devices complete their boot sequence
    \item Ensure that all icons of the devices have a green status icon
\end{enumerate}

\subsubsection{Accessing Device Terminals}

In order to talk to network devices and containers:

\noindent
\\
\textbf{For Cisco Routers:}
\begin{enumerate}
    \item Right-click on any router device
    \item Select \textbf{"Console"} from the context menu
    \item This gives access to the router's command-line interface
\end{enumerate}

\noindent
\textbf{For Docker Containers:}
\begin{enumerate}
    \item Right-click on any container (server or client)
    \item Choose \textbf{"Auxiliary Console"} from the context menu
    \item This opens the container's terminal interface
    \item Use this terminal for all commands and configurations related to containers
\end{enumerate}

\textbf{Key Note:} For Docker containers, always use the "Auxiliary Console" option and not the standard console, as it gives the correct terminal interface to work with the containerized operating system.

\subsection{Router Configuration}

Now that you have access to the router terminals, go ahead and configure the three routers with individual configurations to mimic realistic Internet connectivity and routing behavior.

\subsubsection{ISP Router (R1-Router-ISP)}

The ISP router provides Internet connectivity and inter-site routing. The setup includes interface setup, MAC address spoofing for DHCP, and routing table entries.

\begin{lstlisting}[language=bash]
enable
configure terminal

# Configure interface to Router 2 (Server-side)
interface FastEthernet0/0
  ip address 203.0.113.254 255.255.255.0
  no shutdown
exit

# Configure interface to Router 3 (Client-side)  
interface FastEthernet0/1
  ip address 198.51.100.254 255.255.255.0
  no shutdown
exit

# Configure Internet interface with DHCP and MAC spoofing
interface FastEthernet1/0
  mac-address xxxx.xxxx.xxxx  # Replace with host machine MAC
  ip address dhcp
  no shutdown
exit

# Configure static routes for private networks
ip route 198.51.100.0 255.255.255.0 FastEthernet0/1
ip route 203.0.113.0 255.255.255.0 FastEthernet0/0

# Configure default route to Internet
ip route 0.0.0.0 0.0.0.0 FastEthernet1/0

end
write memory
\end{lstlisting}

\noindent
\textbf{Important Note - MAC Address Spoofing:} 

\noindent
This step is required only needed when your host's DHCP server restricts MAC addresses (e.g., Politecnico di Torino WiFi). In other cases, omit this process or adapt accordingly.

\noindent
The MAC address spoofing is required since the Politecnico di Torino WiFi network's DHCP server uses MAC address filtering to ensure security. Without the use of the host machine's MAC address, the DHCP server will not respond with a valid IP address for the router interface, which denies Internet access for the virtualized network. To know your host machine's MAC address:

\begin{lstlisting}[language=bash]
# Display all network interfaces and their MAC addresses
ip a

# Look for the active network interface (usually wlan0 for WiFi or eth0 for Ethernet)
# The MAC address appears after "link/ether"
# Example output:
# 2: wlan0: <BROADCAST,MULTICAST,UP,LOWER_UP> mtu 1500 qdisc mq state UP group default qlen 1000
#     link/ether 1c:ce:51:3f:98:20 brd ff:ff:ff:ff:ff:ff
\end{lstlisting}

\noindent
Copy the MAC address from your active network interface and convert it to Cisco format by adding dots every four characters:
\begin{itemize}
    \item \textbf{Linux format:} aa:bb:cc:dd:ee:ff
    \item \textbf{Cisco format:} aabb.ccdd.eeff
\end{itemize}

Replace \texttt{xxxx.xxxx.xxxx} in the configuration with your converted MAC address.

\subsubsection{Server-Side Router (R2-Router-Server)}

This router connects the server's private network to the Internet and implements NAT with port forwarding for VPN services.

\begin{lstlisting}[language=bash]
enable
configure terminal

# Configure LAN interface (connected to server)
interface FastEthernet0/0
  ip address 10.0.1.1 255.255.255.0
  ip nat inside
  no shutdown
exit

# Configure WAN interface (connected to ISP)
interface FastEthernet0/1
  ip address 203.0.113.1 255.255.255.0
  ip nat outside
  no shutdown
exit

# Configure static NAT for VPN services
ip nat inside source static udp 10.0.1.2 500 203.0.113.1 500
ip nat inside source static udp 10.0.1.2 4500 203.0.113.1 4500
ip nat inside source static tcp 10.0.1.2 443 203.0.113.1 443

# Configure PAT for general Internet access
access-list 1 permit 10.0.1.0 0.0.0.255
ip nat inside source list 1 interface FastEthernet0/1 overload

# Configure default route
ip route 0.0.0.0 0.0.0.0 FastEthernet0/1

end
write memory
\end{lstlisting}

\noindent
\textbf{Port Forwarding Explanation:}
\begin{itemize}
    \item \textbf{Port 500/UDP:} ISAKMP (Internet Security Association and Key Management Protocol)
    \item \textbf{Port 4500/UDP:} NAT-T (NAT Traversal for IPSec)
    \item \textbf{Port 443/TCP:} HTTPS/TLS for SSL VPN connectivity
\end{itemize}

\subsubsection{Client-Side Router (R3-Router-Client)}

The client-side router provides NAT and routing for the client's private network.

\begin{lstlisting}[language=bash]
enable
configure terminal

# Configure LAN interface (connected to client)
interface FastEthernet0/0
  ip address 10.0.2.1 255.255.255.0
  ip nat inside
  no shutdown
exit

# Configure WAN interface (connected to ISP)
interface FastEthernet0/1
  ip address 198.51.100.1 255.255.255.0
  ip nat outside
  no shutdown
exit

# Configure default route
ip route 0.0.0.0 0.0.0.0 FastEthernet0/1

# Configure PAT for Internet access
access-list 1 permit 10.0.2.0 0.0.0.255
ip nat inside source list 1 interface FastEthernet0/1 overload

end
write memory
\end{lstlisting}

\subsection{Container Network Configuration}

Now that you have access to the container terminals, configure the network interfaces for both containers.

\subsubsection{Server Network Configuration}

Configure the server container's network interface:

\begin{lstlisting}[language=bash]
# Configure IP address and default route
ip addr add 10.0.1.2/24 dev eth0
ip route add default via 10.0.1.1

# Verify network configuration
ip addr show
ip route show

# Test connectivity to gateway
ping 10.0.1.1
\end{lstlisting}

\subsubsection{Client Network Configuration}

Configure the client container's network interface and install required VPN software:

\begin{lstlisting}[language=bash]
# Configure network interface
ip addr add 10.0.2.2/24 dev eth0
ip route add default via 10.0.2.1

# Update package repositories
apt update

# Install strongSwan for IPSec VPN
apt install strongswan -y

# Install OpenVPN for TLS VPN  
apt install openvpn -y

# Copy configuration files from persistent storage
cp /client/ipsec.conf /etc/
cp /client/ipsec.secrets /etc/

# Verify network configuration
ip addr show
ip route show

# Test connectivity to gateway
ping 10.0.2.1
\end{lstlisting}

\subsection{Basic Connectivity Testing}

Before proceeding with VPN configuration, verify that the basic network infrastructure is functioning correctly.

\subsubsection{Inter-Router Connectivity}

Test connectivity between routers to ensure proper routing:

\begin{lstlisting}[language=bash]
# From ISP Router - test connectivity to edge routers
ping 203.0.113.1    # Should reach Server-side router
ping 198.51.100.1   # Should reach Client-side router

# From Server-side Router - test connectivity to ISP
ping 203.0.113.254  # Should reach ISP router

# From Client-side Router - test connectivity to ISP  
ping 198.51.100.254 # Should reach ISP router
\end{lstlisting}

\subsubsection{End-to-End Connectivity}

Test connectivity between the container endpoints:

\begin{lstlisting}[language=bash]
# From Server Container - test connectivity to public IPs
ping 198.51.100.1   # Should reach Client-side router public IP

# From Client Container - test connectivity to public IPs
ping 203.0.113.1    # Should reach Server-side router public IP
\end{lstlisting}

\subsubsection{Service Verification}

Verify that the SoftEther VPN server is running and listening on required ports:

\begin{lstlisting}[language=bash]
# Check if SoftEther VPN server is listening
ss -tuln | grep -E '(443|500|4500)'

\end{lstlisting}

\subsection{Network Infrastructure Verification}

Now, the fundamental network infrastructure must be up and running with:

\begin{itemize}
    \item Three routers configured with appropriate IP addressing and routing
    \item NAT services functioning on edge routers
    \item Docker containers with network connectivity
    \item SoftEther VPN server operational and accessible
    \item Straightforward inter-site connectivity through emulated Internet
\end{itemize}

\noindent
This configuration offers the foundation for deployment of VPN services described in the subsequent sections. Any connectivity issues at this point must be resolved before configuring VPN because they will prevent successful creation of VPN tunnels.

\noindent
The next section will describes how to configure the SoftEther VPN server to provide multi-protocol VPN services for both IPSec and TLS-based connections.
